Discussion Draft

### 2020-04-05 doc/Discussion.rmd
little more

@@

Under the **spindle based** model, faster male evolution of the genome wide recombination rates is primarily driven the strength of the spindle assembly checkpoint (SAC) acting on the chromosome structure at metaphase alignment. We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.

The transition from metaphase to anaphase is irreversible, so prevention of annueploidy requires that all bivalents be correctly aligned before meiosis can progress (cite). Unstable bivalents detected on the MI spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize / connect to the microtubule spindle. If the 'unattached' / unstable bivalents fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents, will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle.

Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis (LaneKauppi, cite). The different the detection of achiasmated bivalents (), the rates of annuploidy in mature gametes, and stronger infertility phenotypes in mutants of key SAC genes (cite).

These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volume (cite).

-- Supporting this stage/cell physiology based mechanism  influencing broad sexual dimorphic phenotypes in recombination landscape. (leading to distinct evolutionary trajectories) **directional selection** in males and relaxed selection in females.

>>>>

**connecting reverse HetC to faster male evolution**

@@

Under the **spindle based** model, faster male evolution of the genome wide recombination rate, (and instances of reversed heterochiasmy) is primarily driven by a more efficient spindle assembly checkpoint (SAC) acting on chromosome structures at metaphase I alignment. <We refer the reader to [@laneKauppi2019] for more details on the SAC pathway.> The SAC prevents anueploidy by ensuring all bivalents are correctly attached to the microtubulue spindle before starting the irreversable metaphase-to-anaphase transition (cite).  A stable chromosome structure requires, i) kinetochores attched to opposite poles of the cells and ii) at least one crossover to create tension across a propotion of sister cohesion (cite). Unstable bivalents on the MI spindle, cause the SAC to delay anaphase and allow time for the bivalent to stabilize on the microtubule spindle. If unstable bivalents fail to attach, the SAC triggers apoptosis. Genetic variants which increase the number of unstable bivalents (such as those affecting recombination rate cohesion), will be selected against in favor for genetic variants causing faster stabilization of bivalents on the MI spindle.

Multiple lines of evidence indicate that the effectiveness of SAC in spermatogenesis is greater compared to oogenesis (LaneKauppi, Sub cite). The different the detection of achiasmated bivalents (), the rates of annuploidy in mature gametes, and stronger infertility phenotypes in mutants of key SAC genes (cite).

These differences in SAC strength have been connected to conserved features of gametogenesis; centrosome spindle and cell volume (cite).
These results // Supporting this stage/cell physiology based mechanism  influencing broad sexual dimorphic phenotypes in recombination landscape. (leading to distinct evolutionary trajectories) **directional selection** in males and relaxed selection in females.
